---
title: "code_for_covid_gr_clinics_with_updates"
author: "Argyris Stringaris"
date: "5/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, results="hide", message=FALSE}
 #install.packages(c("dplyr", "ggplot2", "tidyverse", "patchwork", "tidyr", "rsample",
 #                   "gplots", "lme4", "nlme", "psych", "cowplot", "lavaan", "semPlot", #"semTools", "wBoot", "hrbrthemes", "MuMIn"))

library(dplyr)
# library(data.table)
library(ggplot2)
# #library(SummarizedExperiment)
# #library(ggstatsplot)
# #library(ggpubr)
#library(tidyverse)
library(patchwork)
library(tidyr)
# library(broom)
# library(broom.mixed)
# library(gmodels)
# library(ggthemes)
# library(SuppDists)
# library(questionr)
library(rsample)
# library(zoo)
# library(pracma)
# library(Metrics)
library(gplots)
# library(gmodels)
 library(lme4)
 library(nlme)
# library(kableExtra)
library(psych)
# library(car)
# library(eply)
# library(sjmisc)
library(cowplot)
library(lavaan)
library(semPlot)
library(semTools)
library(hrbrthemes)
library(wBoot)
library(MuMIn)
results = "hide"
```

**import dataset that was sent to me on 26th April.**
```{r}
 # Make sure to change source to yours
covid_gr_clinics <- read.csv("~/Documents/greek_covid/COVIDparents_final_IOANNA_1_KK_AG.csv", sep=";", comment.char="#")
```

**select variables to retain**
```{r}
covid_gr_clinics <- covid_gr_clinics%>%
  select(id, family_id, MOOD_STATES_base, MOOD_STATES_3m, DAILY_BEHAVIORS_3M_KI_SUM,DAILY_BEHAVIORS_BASE_KI_SUM, age_D, site, Diagnosis_FINAL_groups_revised,       
  Diagnosis_FINAL_groups_revised_rec)
```


**Rename the variables to make wrangling easier**
```{r}
covid_gr_clinics <- dplyr::rename(covid_gr_clinics, mood_baseline = MOOD_STATES_base)
covid_gr_clinics <- dplyr::rename(covid_gr_clinics, mood_3m = MOOD_STATES_3m)
covid_gr_clinics <- dplyr::rename(covid_gr_clinics, behaviours_baseline = DAILY_BEHAVIORS_BASE_KI_SUM)
covid_gr_clinics <- dplyr::rename(covid_gr_clinics, behaviours_3m = DAILY_BEHAVIORS_3M_KI_SUM)
```

**Split the dataset into a 60-40 for further work**
```{r}
set.seed(1234)
split_covid_gr_clinics <- initial_split(covid_gr_clinics, 0.6)
covid_gr_clinics_train <- training(split_covid_gr_clinics)
covid_gr_clinics_test <- testing(split_covid_gr_clinics)
```

**Turn wide dataframe to long for plotting**
```{r}


# order to be able to see where you are
covid_gr_clinics_train <- covid_gr_clinics_train[order(covid_gr_clinics_train$id),] 

# this code allows you to turn into long multiple variables--break it down pipe-by-pipe to understand it better.
covid_gr_clinics_train_long <- covid_gr_clinics_train %>% 
  gather (key, val, c(mood_baseline, mood_3m, behaviours_baseline, behaviours_3m)) %>% 
   separate(key, c("type", "time_point")) %>% 
   spread(type, val)
```

**Some code to run multiple plots at once**
```{r}

# first a quick and dirty function to create an exploratory histogram--note [7] is the position of the time_point variable in the dataframe. This may change. It is not ideal this way, will change later
explore_hist <- function(yvar){
    ggplot(covid_gr_clinics_train_long, aes_(x=as.name(yvar))) +
        geom_histogram(aes_string(fill = names(covid_gr_clinics_train_long[7])), position = "dodge") +
        facet_wrap(~age_D) # again, facet wrapped by inputting the variable here--not good practice
}

# this allows you to get the plot over the variables. note [c(8,9)] is the position of the variables of interest in the dataset. They could be in a different position of course or be more
lapply(names(covid_gr_clinics_train_long[c(8:9)]), explore_hist)

# this double lapply allows you to get it over another set of levels, such as site in this example
# the first lapply splits it all by site (or whatever else you wish), then you apply the lapply as above.
lapply(split(covid_gr_clinics_train_long, covid_gr_clinics_train_long$site), function(x) lapply(names(covid_gr_clinics_train_long[c(8:9)]), explore_hist))
```


**A function to get bootstrapped mean differences on a large scale**
```{r}
# first gather the names of the variables you need and turn into a dataframe
three_months <- names(covid_gr_clinics_train %>% select(ends_with("3m")))
baseline <- names(covid_gr_clinics_train %>% select(ends_with("baseline")))

df_3m_baseline <- as.data.frame(cbind(three_months, baseline))

# write the function to extract the bootstrapped means
doing_bootstrap_2 <- function (df, x, y) {
  library(wBoot)
set.seed(1234)
bootstrap_result <-   boot.paired.per(df[, x], df[, y],  null.hyp = NULL,
                 R = 10000)

bootstrap_data <- bootstrap_result[[1]] # 

return(bootstrap_data)
}

# apply the function over each pair (3m - baseline) of variables
library(foreach)
vec_bootstr_results <- as.data.frame(foreach(x = df_3m_baseline[, "three_months"], y = df_3m_baseline[, "baseline"],  .combine='cbind') %do% doing_bootstrap_2(df = covid_gr_clinics_train, x, y))

# rename columns of the bootstrap results
extract_column_names <- gsub("_.*$","",df_3m_baseline[,1])
names(vec_bootstr_results)<-extract_column_names 
#vec_bootstr_results
```


** create a table with means, confidence intervals and p-values from the bootstrap**
```{r}
# create table with results
vec_bootstr_results_means <- sapply(vec_bootstr_results, mean, 2)
vec_bootstr_results_quantiles <- sapply(vec_bootstr_results, quantile, probs = c(0.025, 0.975), 2)
vec_bootstr_results_table <- rbind(vec_bootstr_results_means, vec_bootstr_results_quantiles)
vec_bootstr_results_table <- as.data.frame(as.matrix(t(vec_bootstr_results_table))) # this is to make it easier to read
vec_bootstr_results_table <- vec_bootstr_results_table %>% 
  rename (bootstrap_mean = vec_bootstr_results_means, conf_int_2.5 = "2.5%", conf_int_97.5= "97.5%")
vec_bootstr_results_table
```

** visualise results by creating a separate graph for each column of the database**
```{r}
for (i in names(vec_bootstr_results)){
  plt <-ggplot(vec_bootstr_results, aes_string(x=i))+
         geom_histogram(binwidth = 0.0150) + 
    geom_vline(xintercept = 0, size = 0.5, linetype = 'dashed',colour = "red") +
    geom_vline(xintercept = mean(vec_bootstr_results[[i]]), colour = "blue") +
    geom_vline(xintercept = quantile(vec_bootstr_results[[i]], probs = c(0.025, 0.975)), colour = "black", linetype = "dashed") +
    ggtitle(names(vec_bootstr_results[i]))
  print(plt)
}
```
```{r}
explore_hist <- function (df  , x  , fill  , facet  )
for (i in names(covid_gr_clinics_train_long[8:9])){
  plt <-ggplot(covid_gr_clinics_train_long, aes_string(x=i))+
         geom_histogram(aes_string(fill = names(covid_gr_clinics_train_long[7])), position = "dodge")+
    facet_wrap(~age_D)+
    ggtitle(names(covid_gr_clinics_train_long[i]))
  print(plt)
}
```

**a function to explore the variables in the dataset**
```{r}
explore_hist <- function (df, indices, fill){
for (i in names(df[indices])){
  plt <-ggplot(df, aes_string(x=i))+
         geom_histogram(aes_string(fill = fill), position = "dodge")+
          facet_wrap (~ age_D)+ # this is annoying, it won't accept an argument for some reason
           ggtitle(names(df[i]))
  print(plt)
}
}
explore_hist(covid_gr_clinics_train_long, indices=c(8:9), fill = "time_point")


```
**Another nice way to visualise mood the pre-post**
```{r}
if (!require(remotes)) {
  install.packages("remotes")
}
remotes::install_github('jorvlan/raincloudplots')

library(raincloudplots)

provisional_baseline_mood <- covid_gr_clinics_train_long %>% 
  filter(time_point == "baseline") %>% 
  select (mood, time_point) 

provisional_baseline_mood$time_point <- recode(provisional_baseline_mood$time_point, "baseline" = 0)

provisional_3m_mood <- covid_gr_clinics_train_long %>% 
  filter(time_point == "3m") %>% 
  select (mood, time_point)

provisional_3m_mood$time_point <- recode(provisional_3m_mood$time_point, "3m" = 1)



provisional_baseline_behaviours <- covid_gr_clinics_train_long %>% 
  filter(time_point == "baseline") %>% 
  select (behaviours, time_point) 

provisional_baseline_behaviours$time_point <- recode(provisional_baseline_behaviours$time_point, "baseline" = 0)

provisional_3m_behaviours <- covid_gr_clinics_train_long %>% 
  filter(time_point == "3m") %>% 
  select (behaviours, time_point)

provisional_3m_behaviours$time_point <- recode(provisional_3m_behaviours$time_point, "3m" = 1)





df_1x1_mood <- data_1x1(
  array_1 = provisional_3m_mood[,1], 
  array_2 = provisional_baseline_mood[,1],
  jit_distance = 0.2,
  jit_seed = 321)




raincloud_2_aligned_mood <- raincloud_1x1_repmes(
  data = df_1x1_mood,
  colors = (c('dodgerblue', 'darkorange')),
  fills = (c('dodgerblue', 'darkorange')),
  line_color = 'gray',
  line_alpha = .3,
  size = 1,
  alpha = .6,
  align_clouds = TRUE) +
  
  scale_x_continuous(breaks=c(1,2), labels=c("3 m Before Pandemic", "Baseline"), limits=c(0, 3)) +
  xlab("Time Point") + 
  ylab("Mood") +
  theme_classic()

raincloud_2_aligned_mood

df_1x1_behaviours <- data_1x1(
  array_1 = provisional_3m_behaviours[,1], 
  array_2 = provisional_baseline_behaviours[,1],
  jit_distance = 0.2,
  jit_seed = 321)




raincloud_2_aligned_behaviours <- raincloud_1x1_repmes(
  data = df_1x1_behaviours,
  colors = (c('dodgerblue', 'darkorange')),
  fills = (c('dodgerblue', 'darkorange')),
  line_color = 'gray',
  line_alpha = .3,
  size = 1,
  alpha = .6,
  align_clouds = TRUE) +
  
  scale_x_continuous(breaks=c(1,2), labels=c("3 m Before Pandemic", "Baseline"), limits=c(0, 3)) +
  xlab("Time Point") + 
  ylab("Behaviours") +
  theme_classic()

raincloud_2_aligned_behaviours



library(patchwork)
print(raincloud_2_aligned_behaviours / raincloud_2_aligned_mood)
```

